---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
[![Travis build status](https://travis-ci.org/llrs/BaseSet.svg?branch=master)](https://travis-ci.org/llrs/BaseSet)
[![Coverage status](https://codecov.io/gh/llrs/BaseSet/branch/master/graph/badge.svg)](https://codecov.io/github/llrs/BaseSet?branch=master)
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)

# BaseSet

The goal of BaseSet is to facilitate working with sets in an efficient way.

## Installation

You can install the latest version of BaseSet from [Github](https://github.com/llrs/BaseSet) with:

```{r eval=FALSE}
remotes::install_github("llrs/BaseSet")
```
 
## Examples

```{r include=FALSE}
library("BaseSet")
```

To work with sets we need to have the relations between the elements and the sets:

```{r TidySet}
sets <- list(A = letters[1:5], B = c("a", "f"))
sets_analysis <- tidySet(sets)
```

Perform typical operations like union, intersection:

```{r union-intersection}
union(sets_analysis, sets = c("A", "B")) 
# Or we can give a name to the new set and keep the others sets
union(sets_analysis, sets = c("A", "B"), name = "D")
# We can do the same in the intersection
intersection(sets_analysis, sets = c("A", "B"), name = "D") 
# Or we can omit the new name: 
intersection(sets_analysis, sets = c("A", "B"))
```

And compute size of sets among other things:

```{r set_size}
set_size(sets_analysis)
```

The elements in one set not present in other:

```{r subraction}
subtract(sets_analysis, set_in = "A", not_in = "B", keep = FALSE)
```

Or any other verb from [dplyr](https://cran.r-project.org/package=dplyr). Like mutate

```{r dplyr}
library("magrittr")
set.seed(4673) # To make it reproducible in your machine
sets_analysis %>% 
  mutate(Keep = sample(c(TRUE, FALSE), 7, replace = TRUE)) %>% 
  filter(Keep == TRUE) %>% 
  activate("sets") %>% 
  mutate(sets_origin = c("Reactome", "KEGG"))
```

## Fuzzy sets

You can do the same operations with fuzzy sets:

```{r fuzzy}
relations <- data.frame(sets = c(rep("A", 5), "B", "B"), 
                        elements = c("a", "b", "c", "d", "e", "a", "f"),
                        fuzzy = runif(7))
fuzzy_set <- tidySet(relations)
fuzzy_set
union(fuzzy_set, sets = c("A", "B"))
union(fuzzy_set, sets = c("A", "B"), name = "D", keep = TRUE)
intersection(fuzzy_set, sets = c("A", "B"), name = "D") 
intersection(fuzzy_set, sets = c("A", "B"), keep = TRUE)
# Note here the difference:
set_size(fuzzy_set) # A set could be empty!
element_size(fuzzy_set)
fuzzy_set %>% 
  mutate(Keep = ifelse(fuzzy > 0.5, TRUE, FALSE)) %>% 
  filter(Keep == TRUE) %>% 
  activate("sets") %>% 
  mutate(sets_origin = c("Reactome", "KEGG"))
```

