---
title: "Advanced examples"
abstract: >
  This vignette assumes you are familiar with set operations from the basic vignette.
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('BaseSet')`"
output:
  BiocStyle::html_document:
    fig_caption: true
    code_folding: show
    self_contained: yes
    toc_float:
      collapsed: true
      toc_depth: 3
author:
- name: Lluís Revilla
  affiliation: 
    - August Pi i Sunyer Biomedical Research Institute (IDIBAPS); Liver Unit, Hospital Clinic
  email: lluis.revilla@gmail.com
vignette: >
  %\VignetteIndexEntry{2. Advanced examples}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, 
                      warning = TRUE,
                      comment = "#>")
```
To show compatibility with tidy workflows we will use magrittr pipe operator
```{r setupr}
library("dplyr")
library("BaseSet")
```
We can create a tidySet with some sets being the intersection and the union of others:
```{r}
relations <- data.frame(sets = c(rep("A", 5), "B"),
                        elements = letters[seq_len(6)],
                        fuzzy = runif(6))
sample_set <- tidySet(relations) %>% 
  union(c("A", "B"), "C", keep = TRUE) %>% 
  intersection(c("B", "C"), "D", keep = TRUE)

# Number of elements per set
sample_set %>% 
  relations() %>% 
  group_by(sets) %>% 
  count()

# size of the sets and probability
sample_set %>% 
  set_size()
```

# Human GO

We will explore the genes related with GO. 

```{r prepare_GO}
# We load some libraries
library("org.Hs.eg.db")
library("GO.db")
library("ggplot2")

# Prepare the data 
h2GO_TS <- tidySet(org.Hs.egGO)
h2GO <- as.data.frame(org.Hs.egGO)
```
We can now explore if there are differences in evidence usage for each ontology in GO:
```{r evidence_ontology}
library("forcats")
h2GO %>% 
    group_by(Evidence, Ontology) %>% 
    count(name = "Freq") %>% 
    ungroup() %>% 
    mutate(Evidence = fct_reorder2(Evidence, Ontology, -Freq)) %>% 
    ggplot() +
    geom_col(aes(Evidence, Freq)) +
    facet_grid(~Ontology) + 
    theme_bw() +
    coord_flip() +
    xlab("Evidence")
```
We can see that 
This graph doesn't consider that some relationships are better annotated than other:
```{r nEvidence_plot}
h2GO_TS %>% 
  relations() %>% 
  ggplot() +
    geom_histogram(aes(nEvidence), binwidth = 1) +
    theme_minimal() +
    # scale_y_log10() +
    labs(x = "Evidence codes", 
         title = "Evidence codes for each relation between GO and a gene",
         subtitle = "in human annotation")
```
We can see that mostly all the annotations are done with a single evidence code.
So far we have explored the code that it is related to a gene but how many genes don't have any annotation?
```{r numbers}
# Add all the genes and GO terms
library("GO.db")
h2GO_TS <- add_elements(h2GO_TS, keys(org.Hs.eg.db)) %>% 
  add_sets(grep("^GO:", keys(GO.db), value = TRUE))

sizes_element <- element_size(h2GO_TS) %>% 
    arrange(desc(size))
sum(sizes_element$size == 0)
sum(sizes_element$size != 0)

sizes_set <- set_size(h2GO_TS) %>% 
    arrange(desc(size))
sum(sizes_set$size == 0)
sum(sizes_set$size != 0)
```
So we can see that both there are more genes without annotation and more GO terms without a (direct) gene annotated.

```{r plots_GO}
sizes_element %>% 
    filter(size != 0) %>% 
    ggplot() +
    geom_histogram(aes(size), binwidth = 1) +
    theme_minimal() +
    labs(x = "# sets per element", y = "Count")

sizes_set %>% 
    filter(size != 0) %>% 
    ggplot() +
    geom_histogram(aes(size), binwidth = 1) +
    theme_minimal() +
    labs(x = "# elements per set", y = "Count")
```
As you can see on the second plot we have very large values but that are on associated on many genes:
```{r distr_sizes}
head(sizes_set, 10)
```

# Human pathways

Now we will repeat the same analysis with pathways:

```{r prepare_reactome}
# We load some libraries
library("reactome.db")

# Prepare the data (is easier, there isn't any ontoogy or evidence column)
h2p <- as.data.frame(reactomeEXTID2PATHID)
colnames(h2p) <- c("sets", "elements")
# Filter only for human pathways
h2p <- h2p[grepl("^R-HSA-", h2p$sets), ]

# There are duplicate relations with different evidence codes!!: 
summary(duplicated(h2p[, c("elements", "sets")]))
h2p <- unique(h2p)
# Create a tidySet and 
h2p_TS <- tidySet(h2p) %>% 
    # Add all the genes 
    add_elements(keys(org.Hs.eg.db))
```
Now that we have everything ready we can start measuring some things..
```{r numbers_pathways}
sizes_element <- element_size(h2p_TS) %>% 
    arrange(desc(size))
sum(sizes_element$size == 0)
sum(sizes_element$size != 0)

sizes_set <- set_size(h2p_TS) %>% 
    arrange(desc(size))
```
We can see there are more genes without pathways than genes with pathways.

```{r pathways_plots}
sizes_element %>% 
    filter(size != 0) %>% 
    ggplot() +
    geom_histogram(aes(size), binwidth = 1) +
    scale_y_log10() +
    theme_minimal() +
    labs(x = "# sets per element", y = "Count")

sizes_set %>% 
    ggplot() +
    geom_histogram(aes(size), binwidth = 1) +
    scale_y_log10() +
    theme_minimal() +
    labs(x = "# elements per set", y = "Count")
```
As you can see on the second plot we have very large values but that are on associated on many genes:
```{r distr_sizes_pathways}
head(sizes_set, 10)
```

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
