---
title: "About BaseSet"
abstract: >
  Describes the background of the package, important functions defined in the
  package and some of the applications and usages.
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('BaseSet')`"
output:
  BiocStyle::html_document:
    fig_caption: true
    code_folding: show
    self_contained: yes
    toc_float:
      collapsed: true
      toc_depth: 3
author:
- name: Lluís Revilla
  affiliation: 
    - August Pi i Sunyer Biomedical Research Institute (IDIBAPS); Liver Unit, Hospital Clinic
  email: lluis.revilla@gmail.com
vignette: >
  %\VignetteIndexEntry{1. Introduction}
  %\VignetteEncoding{UTF-8}  
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(collapse = TRUE, 
                      warning = TRUE,
                      comment = "#>")
BiocStyle::markdown()
library("BiocStyle")
```

# Getting started

This vignette explains how to use the methods available in this package.


# The TidySet class

This is a basic example which shows you how to create a `TidySet` object, to store associations between genes and sets:

```{r from_list, message=FALSE}
library("BaseSet")
gene_lists <- list(
    geneset1 = c("A", "B"),
    geneset2 = c("B", "C", "D")
)
tidy_set <- tidySet(gene_lists)
tidy_set
```

This is then stored internaly in three slots `relations`, `elements`, and `sets` slots.

If you have more information for each element or set it can be used added:

```{r metadata, message=FALSE}
gene_data <- data.frame(
    stat1     = c( 1,   2,   3,   4 ),
    info1     = c("a", "b", "c", "d")
)

tidy_set <- add_column(tidy_set, "elements", gene_data)
set_data <- data.frame(
    Group     = c(      100,        200 ),
    Colum     = c(     "abc",      "def")
)
tidy_set <- add_column(tidy_set, "sets", set_data)
tidy_set
```

This data is stored in one of the three slots, which can be directly accessed using their getter methods:

```{r getters}
relations(tidy_set)
elements(tidy_set)
sets(tidy_set)
```

You can add as much information as you want, with the only restriction of for a "fuzzy" column for the `relations`. This column is restricted to a numeric value between 0 and 1. To indicate the strength (or probability) of the relationship between the element and the set. 

```{r fuzzy}
set.seed(4567) # To be able to have exact replicates
relations <- data.frame(sets = c(rep("A", 5), "B", "C"),
                          elements = c(letters[seq_len(6)], letters[6]),
                          fuzzy = runif(7))
fuzzy_set <- tidySet(relations)
```

# Creating a TidySet

As you can see it is possible to create a TidySet from a list and a data.frame, but it is also possible from a matrix:

```{r tidySet_matrix}
m <- matrix(c(0, 0, 1, 1, 1, 1), ncol = 2, nrow =3,  
               dimnames = list(letters[1:3], LETTERS[1:2]))
m
tidySet(m)
```

Or they can be created from a GeneSet and GeneSetCollection objects. Additionally it has several function to read files related to sets like the OBO files (`getOBO`) and GAF (`getGAF`)

# Converting to other formats

It is possible to extract the gene sets as a `list`, for use with functions such as `lapply`.

```{r as.list}
as.list(tidy_set)
as.list(fuzzy_set)
```

Or if you need to apply some network methods and you need a matrix, you can create it with `incidence`:

```{r incidence}
incidence(tidy_set)
incidence(fuzzy_set)
```

# Operations with sets

To work with sets several methods are provided. In general you can provide a new name for the resulting set of the operation, but if you don't one will be automatically provided using `naming`. All methods work with fuzzy and non-fuzzy sets

## Union

You can make a union of two sets present on the same object.

```{r union, eval=FALSE}
union(fuzzy_set, sets = c("C", "B"), name = "D")
union(fuzzy_set, sets = c("C", "B"), name =  "D", FUN = "mean")
```
```{r false_union, eval=TRUE, echo=FALSE, results="show"}
BaseSet::union(fuzzy_set, sets = c("C", "B"), name = "D")
BaseSet::union(fuzzy_set, sets = c("C", "B"), name =  "D", FUN = "mean")
```

Observe tha the fuzziness on the resulting set has changed (on the traditional sets it doesn't matter)

## Intersection


```{r intersection}
intersection(fuzzy_set, sets = c("A", "B"), name = "D", keep = TRUE)
```

The keep argument used here is if you want to keep all the other previous sets:

```{r intersection}
intersection(fuzzy_set, sets = c("A", "B"), name = "D", keep = FALSE)
```

## Complement

We can look for the complement of one or several sets:

```{r complement}
complement_set(fuzzy_set, sets = c("A", "B"))
```

Observe that we haven't provided a name for the resulting set but we can provide one if we prefer to

```{r complement}
complement_set(fuzzy_set, sets = c("A", "B"), name = "F")
```

## Subtract

This is the equivalent of `setdiff`, but clearer:

```{r subtract}
subtract(fuzzy_set, set_in = "A", not_in = "B", keep = FALSE, name = "A-B")
subtract(fuzzy_set, set_in = "B", not_in = "A", keep = FALSE)
```


# Additional information

The number of unique elements and sets can be obtained using the `nElements` and `nSets` methods.

```{r n}
nElements(tidy_set)
nSets(tidy_set)
nRelations(tidy_set)
```

The size of each gene set can be obtained using the `set_size` method.

```{r set_size}
set_size(tidy_set, "geneset1")
set_size(fuzzy_set, "A")
```

Conversely, the number of sets associated with each gene is returned by the `elementLengths` function.

```{r element_size}
element_size(tidy_set)
```

The identifiers of elements and sets can be inspected and renamed using `name_elements` and 

```{r name}
name_elements(tidy_set)
name_elements(tidy_set) <- paste0("Gene", seq_len(nElements(tidy_set)))
name_elements(tidy_set)
name_sets(tidy_set)
name_sets(tidy_set) <- paste0("Geneset", seq_len(nSets(tidy_set)))
name_sets(tidy_set)
```


# Using `dplyr` verbs

You can also use `mutate`, `filter` and other `dplyr` verbs with TidySets (with the only exception being `group_by`), but you usually need to activate which three slots you want to affect with `activate`:

```{r tidyverse}
library("dplyr")
fuzzy_set %>% 
  activate("relations") %>% 
  mutate(Important = as.numeric(fuzzy > 0.5))
```

You can use activate to select what are the verbs modifying:

```{r deactivate}
set_modified <- fuzzy_set %>% 
  activate("relations") %>% 
  mutate(Important = as.numeric(fuzzy > 0.5)) %>% 
  activate("elements") %>% 
  mutate(Pathway = if_else(elements %in% c("b", "f", "g"), 
                           "pathway1", 
                           "pathway2"))
set_modified
set_modified %>% 
  deactivate() %>% # To apply a filter independently to where it is
  filter(Pathway == "pathway1", Important == 1)
```


If you think you need group_by usually this would mean that you need a new set. You can create a new one with `group`. 
If you want to use `group_by` to group some elements then you need to create a 
new set:

```{r group}
# A new group of those elements in pathway1 and with Important == 1
set_modified %>% 
  deactivate() %>% 
  group(name = "new", Pathway == "pathway1", Important == 1)
```
```{r group}
fuzzy_set %>% 
  group("pathway1", elements %in% c("b", "f", "g"))
```

After grouping or mutating sometimes we might be interested in moving a column describing something to other places. We can do by this with:

```{r moving}
elements(set_modified)
out <- move_to(set_modified, "elements", "relations", "Pathway")
relations(out)
```

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
