---
title: "About BaseSet"
abstract: >
  Describes the background of the package, important functions defined in the
  package and some of the applications and usages.
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('BaseSet')`"
output:
  BiocStyle::html_document:
    fig_caption: true
    code_folding: show
    self_contained: yes
    toc_float:
      
      collapsed: true
      toc_depth: 3
author:
- name: Lluís Revilla
  affiliation: 
    - August Pi i Sunyer Biomedical Research Institute (IDIBAPS); Liver Unit, Hospital Clinic
  email: lluis.revilla@gmail.com
vignette: >
  %\VignetteIndexEntry{About BaseSet}
  %\VignetteEncoding{UTF-8}  
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
```{r knitsetup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(collapse = TRUE, warning = TRUE)
BiocStyle::markdown()
library("BiocStyle")
```

# Getting started

This vignette explains how to use the methods available in this package.


## The BaseSets class

This is a basic example which shows you how to create a `TidySet` object, to store associations between genes and sets:

```{r from_list, message=FALSE}
library("BaseSet")
gene_lists <- list(
    geneset1 = c("A", "B"),
    geneset2 = c("B", "C", "D")
)
tidy_set <- tidySet(gene_lists)
tidy_set
```

Metadata for each element and set can be added:

```{r metadata, message=FALSE}
gene_data <- data.frame(
    stat1     = c( 1,   2,   3,   4 ),
    info1     = c("a", "b", "c", "d")
)

tidy_set <- add_column(tidy_set, "elements", gene_data)
set_data <- data.frame(
    Group     = c(      100,        200 ),
    Colum     = c(     "abc",      "def")
)
tidy_set <- add_column(tidy_set, "sets", set_data)
tidy_set
```

The `relations`, `elements`, and `sets` slots can directly accessed using their getter methods.

```{r getters}
relations(tidy_set)
elements(tidy_set)
sets(tidy_set)
```

All the relations are inherently fuzzy

```{r fuzzy}
relations <- data.frame(sets = c(rep("A", 5), "B", "C"),
                          elements = c(letters[seq_len(6)], letters[6]),
                          fuzzy = runif(7))
fuzzy_set <- tidySet(relations)
```


# Converting to other formats

It is possible to extract the gene sets as a `list`, for use with functions such as `lapply`.

```{r as.list}
(list_set <- as.list(tidy_set))
(list_set <- as.list(fuzzy_set))
```

It is also possible to visualize membership between gene and gene sets as a matrix.

Notably, `incidence` return a `numeric` matrix of membership:

```{r incidence}
(incidence_matrix <- incidence(tidy_set))
(incidence_matrix <- incidence(fuzzy_set))
```


# Converting from other formats

Conversely, it is also possible to make objects from incidence matrices.

Notably, `BaseSets` are suitable for `numeric` matrices indicating membership:

```{r tidySet_matrix}
tidySet(incidence_matrix)
```

Or can also be created from lists:
```{r tidySet_list}
tidySet(list("A" = letters[1:4], "B" = letters[7:10]))
```

# Additional information

The number of unique elements and sets can be obtained using the `nElements` and `nSets` methods.

```{r n}
nElements(tidy_set)
nSets(tidy_set)
nRelations(tidy_set)
```

The size of each gene set can be obtained using the `set_size` method.

```{r set_size}
set_size(tidy_set, "geneset1")
set_size(fuzzy_set, "A")
```

Conversely, the number of sets associated with each gene is returned by the `elementLengths` function.

```{r element_size}
element_size(tidy_set)
```

The identifiers of elements and sets can be inspected and renamed using `name_elements` and 

```{r name}
name_elements(tidy_set)
name_elements(tidy_set) <- paste0("Gene", seq_len(nElements(tidy_set)))
name_elements(tidy_set)
name_sets(tidy_set)
name_sets(tidy_set) <- paste0("Geneset", seq_len(nSets(tidy_set)))
name_sets(tidy_set)
```

# Operations with sets

To work with sets several methods are provided. All methods work with fuzzy and non-fuzzy sets

## Union

By default it provides the standard union, but you can provide a function.
```{r union}
union(fuzzy_set, c("C", "B"), "D")
union(fuzzy_set, c("C", "B"), "D", FUN = "mean")
```

## Intersection


```{r intersection}
intersection(fuzzy_set, c("A", "B"), "D", keep = TRUE)
```

## Complement

We can look for the complement of one or several sets:
```{r complement}
complement_set(fuzzy_set, c("A", "B"))
```

# tidyverse

You can also use `mutate`, `filter` and other `dplyr` verbs with TidySets (with the only exception being `group_by`):

```{r tidyverse}
library("dplyr")
fuzzy_set %>% 
  activate("relations") %>% 
  mutate(Important = as.numeric(fuzzy > 0.5))
```

You can use activate to select what are the verbs modifying:

```{r deactivate}
set_modified <- fuzzy_set %>% 
  activate("relations") %>% 
  mutate(Important = as.numeric(fuzzy > 0.5)) %>% 
  activate("elements") %>% 
  mutate(Pathway = if_else(elements %in% c("b", "f", "g"), 
                           "pathway1", 
                           "pathway2"))
set_modified
set_modified %>% 
  deactivate() %>% # To apply a filter independently to where it is
  filter(Pathway == "pathway1", Important == 1)
```

If you want to use `group_by` to group some elements then you need to create a 
new set:

```{r group}
# A new group of those elements in pathway1 and with Important == 1
set_modified %>% 
  deactivate() %>% 
  group("new", Pathway == "pathway1", Important == 1)
```

After grouping or mutating sometimes we might be interested in moving a column describing something to other places. We can do by simiply:

```{r}
elements(set_modified)
out <- move_to(set_modified, "elements", "relations", "Pathway")
relations(out)
```

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
